### 2012-11-20 paper/isca40format/Introduction.tex
stuff

Large shared-memory multicores are likely to use cache-coherence protocols
which are directory-based, hierarchical, and implemented using message passing.
Even though such protocols have been around for two decades (see, for example,
DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
such protocols remains a challenge. It is difficult to convince oneself that
all corner cases have been handled adequately without a formal proof. And yet
formal proofs of real implementations are practically non-existent. One can't
settle for conservative choices in cache-coherence protocol designs because
performance (\eg the number of messages needed to maintain coherence, the
buffer requirements, etc) is as important as correctness. Consequently proving
a simple protocol for a small number of caches, under unbounded buffer
assumption is of limited use in gaining confidence in the correctness of a real
protocol.

In this paper we will consider systems where caches form a tree hierarchy and
caches are \emph{inclusive}, that is, if a cache contains address $a$ then its
parent is guaranteed to contain address $a$, though the parent may have stale
data for $a$. Every cache or node in the system has an associated directory
which contains information about its children caches. The coherence state
changes are orchestrated by passing requests and responses between the caches
via the directory.  Unlike bus-based protocols which lock the bus till all the
required state transitions are done thereby ensuring atomicity of state
transitions, in a message-passing protocol, requests and responses are split
leading to transient states. This creates scenarios not present in bus-based
protocols, which can lead to coherence violations or deadlocks, if the protocol
is not designed or implemented carefully.

@@

\begin{enumerate} \item Lots of concurrent events must be dealt with without a
central authority to impose a sequence on events; \item The state on which
transitions are based is physically distributed in space and cannot be accessed
atomically; \item Physical network may block the transmission of a message
because of lack of communication buffers or head-of-the-line blocking; \item
The cache or directory may reach a state which can not handle a particular
message. An incomplete set of transition rules may cause the protocol to
deadlock.  \end{enumerate}

@@

These examples give a glimpse of the complexity of designing and implementing a
correct cache coherence protocol in a directory-based non-bus setting.  Even
for a 2-level cache hierarchy, designing and implementing the protocol is quite
difficult. This problem is greatly exacerbated in a multi-level cache
hierarchy, which is becoming the norm because of the increasing core counts.

In this paper, we present a substantially easier method to design cache
coherence protocols for such systems. The user specifies a protocol for a
two-level system without message passing. The protocol rules can simultaneously
or \emph{atomically} examine and change the state of all the caches in the
system. For example, the MSI protocol may be described as follows:

***** Murli insert the description of MSI **** We give a procedure that
transforms such simple and global protocol descriptions into a distributed,
message-passing protocol for hierarchal caches. Our procedure is based on
preserving some invariants in rule transformations and has been shown to be
sound and complete with respect the global description using the coq theorem
prover \cite{}, however, a discussion of the formal proof is beyond the scope
of this paper. Both the procedure and the proof are parameterized by the number
of cache levels and the number of caches in each level of the hierarchy as well
as by a predicate which involves a local ``compatibility of state'' test.

It should be noted that the flexibility of our procedure comes from the fact
that we do not assign any meaning to the value of a coherence state beyond
upgrades or downgrades. Thus, how or why a particular cache state implies that
a particular address may not be present in any other cache stems from protocol
specific ``compatibility of state'' check.

The main contributions of this paper are: 1. An automatic procedure to refine
centralized protocols without message passing into protocols for distributed,
hierarchal, cache system with message-passing; 2. Specification of the
invariants to be preserved by the procedure; and 3. Showing that our procedure
does not introduce any overhead over direct implementation of these protocols.
We think that our method of designing protocols substantially reduces the
required effort, and that our procedure is flexible enough to implement (while
ensuring correctness of) most, if not all, inclusive message-passing
directory-based protocols.

\paragraph{Paper organization:} *** to be redone *** Section \ref{sec:abstract}
describes the abstract protocol; Section \ref{sec:properties} gives the two
properties that are guaranteed by the abstract protocol and provides some
intuition for proving these properties. We also discuss the communication
buffer requirements for the abstract protocol and show that the requirement is
only proportional to the number of levels and not to the number of processors.
In Section \ref{sec:protocols}, we show how the abstract protocol can be used
to implement MSI, MESI, MOSI and MOESI protocols and their variants.  In
Section \ref{sec:performance}, we show that these formally verified family of
protocols require no extra state bits, nor extra message transfers compared to
other published implementations.  In Section \ref{sec:related}, we present some
of the related work and finally in Section \ref{sec:conclusion}, we present
some directions for the future work.

>>>>

Large cache-coherent shared-memory multicores are likely to use protocols which
are directory-based, hierarchical, and implemented using message passing.  Even
though such protocols have been around for two decades (see, for example,
DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
such protocols remains a challenge. It is difficult to convince oneself that
all corner cases have been handled adequately without a formal proof. And yet
formal proofs of real implementations are practically non-existent. One can't
settle for conservative choices in cache-coherence protocol designs because
performance (\eg the number of messages needed to maintain coherence, the
buffer requirements, etc) is as important as correctness. Consequently proving
a simple protocol for a small number of caches, under unbounded buffer
assumption is of limited use in gaining confidence in the correctness of a real
protocol.

In this paper we will consider systems where caches logically form a tree
hierarchy and caches are \emph{inclusive}, that is, if a cache contains address
$a$ then its parent is guaranteed to contain address $a$, though the parent may
have stale data for $a$ (Figure \ref{hierarchy}). Physically the system looks
like a collection of caches connected to a message passing network (Figure
\ref{physical}). Caches typically contain data and state for each cache line,
and have two pairs of channels for communicating with parent and children
caches. For the lowest level caches (L1's) the channels for communicating with
children are connected to processors, while for the last-level caches the
channels for communicating with the parents are connected to the memory.

Every cache, unless it is an L1 cache, has a directory which contains
information about the state of its children caches. The cache state changes to
maintain coherence are orchestrated by passing requests and responses between
the caches using the local directory information.  Unlike bus-based protocols
which lock the bus till all the required state transitions are done thereby
ensuring atomicity of state transitions, in a message-passing protocol,
requests and responses are split leading to transient states. This creates
scenarios not present in bus-based protocols, which can lead to coherence
violations or deadlocks, if the protocol is not designed or implemented
carefully.

@@

\begin{enumerate} \item Lots of concurrent events must be dealt with without a
central authority to impose a sequence on events; \item The state on which
transitions are based conceptually is physically distributed and cannot be
accessed atomically; \item Physical network may block the transmission of a
message because of lack of communication buffers or head-of-the-line blocking;
\item The cache or directory may reach a state in which a particular message
cannot be handled. An incomplete set of transition rules may cause the protocol
to deadlock.  \end{enumerate}

@@

These examples give a glimpse of the complexity of designing and implementing a
correct cache coherence protocol in a distributed setting.  Even for a 2-level
cache hierarchy, designing and implementing the protocol is non-trivial; the
problem gets greatly exacerbated in large multi-level cache system.

In this paper, we present a substantially easier method to design cache
coherence protocols for such systems. The user specifies a protocol for a cache
hierarchy as sequential methods to process requests. There is no message
passing so the protocol procedure can simultaneously or \emph{atomically}
examine and change the state of all the caches in the system. We then
systematically transform such a protocol description into a distributed,
message-passing protocol. The resulting protocol is presented as a collection
of cooperating sequential methods which operate on ``cache objects''.  These
sequential methods communicate with each other via blocking \emph{send} and
\emph{receive} commands and thus can get suspended. Our transformation is based
on preserving some invariants which guarantee that each request eventually gets
a response and thus, our threads can never wait indefinitely.  It has been
shown using the Coq theorem prover \cite{}, that the distributed protocol we
derive is correct.  Both the transformation procedure and the proof are
parameterized by the number of cache levels and the number of caches in each
level of the hierarchy as well as by a predicate which involves a local
``compatibility of state'' test. A discussion of the formal proof is beyond the
scope of this paper.

It should be noted that the generality of our procedure comes from the fact
that we do not assign any meaning to a coherence state beyond upgrades and
downgrades. Thus, how or why a particular cache state implies that a particular
address may not be present in any other cache stems from protocol specific
``compatibility of state'' check.

The main contributions of this paper are: 1. An automatic procedure to refine a
centralized MSI protocol without message passing into protocols for
distributed, hierarchical, cache system with message-passing; 2. Specification
of the invariants that ensure the correctness of the transformation procedure;
and 3. Giving a provable lower bound on network virtual channels and buffers to
avoid deadlocks; and 4. Showing that our procedure does not introduce any
overhead over direct implementation of these protocols.  We think that our
method of designing protocols substantially reduces the required effort in
implementing provably correct protocols.

\paragraph{Paper organization:} Section \ref{sec:GlobalMsi} describes the MSI
protocol for a hierarchy of caches using nonsuspensive sequential methods. We
use it a specification for proving the correctness of distributed MSI
protocols. Section \ref{sec:DistributedMsi} systematically transforms the
protocol in Section \ref{sec:GlobalMsi} into a distributed protocol. Section
\ref{sec:properties} gives a set of global properties or invariants that are
needed to preserve the correctness of the transformation procedure. The
desirability of these invariants is intuitively obvious. Section
\ref{sec:localProperties} describes local properties that are needed to
guarantee the desirable invariants discussed in Section \ref{sec:properties};
it can be easily seen by inspection (because of the local nature of the
properties) that the distributed protocol obeys these properties. Section
\ref{sec:network} specifies the minimum virtual channel requirements to ensure
the local properties specified in Section \ref{sec:localProperties}. We show
that these formally verified family of protocols require no extra state bits,
nor extra message transfers compared to other published implementations.  In
Section \ref{sec:related}, we present some of the related work and finally in
Section \ref{sec:conclusion}, we present some directions for the future work.

### 2012-11-20 paper/isca40format/Invariants.tex
stuff

\section{Invariants}
\label{sec:invariants}

@@

@@

\begin{figure}\small
\begin{inv}
A response \Resp{p}{c}{a}{x}, where $p = c.parent$, can be sent from $p$ only
if
\begin{enumerate}
\item a thread that handling a request from $c$ to $p$ for
address $a$ is executing, and
\item $x > p.dir[c][a]$
\end{enumerate}
\label{cSendRespPre}
\end{inv}
\begin{inv}
If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$
\label{cSendRespPost}
\end{inv}
\begin{inv}
A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
if $x < c.state[a]$ \label{pSendRespPre1}
\end{inv}
\begin{inv}
If there is a thread in node $c$ that is waiting to receive a response from its
parent $p$ for an address $a$, then response \Resp{c}{p}{a}{x} can be sent only
on executing a thread that is handling a request from $p$ to $c$ for address
$a$. For instance, $a$ can not be evicted from $c$, if there is a thread
handling a request from $p$ to $c$ for address $a$.
\label{pSendRespPre2}
\end{inv}
\begin{inv}
If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$
\label{pSendRespPost}
\end{inv}
\caption{Invariants for sending responses}
\label{sendResp}
\end{figure}

\begin{figure}\small
\begin{inv}
On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
\gets x$
\label{pRecvResp}
\end{inv}
\begin{inv}
On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
\gets x$
\label{cRecvResp}
\end{inv}
\caption{Invariants for receiving responses}
\label{recvResp}
\end{figure}

\begin{figure}\small
\begin{inv}
$c.state[a]$ can change only on $c$ sending a response to or receiving a
response from $c.parent$
\label{cState}
\end{inv}
\begin{inv}
$p.dir[c][a]$ can change only on $p$ sending or receiving a response from
its child $c$
\label{pState}
\end{inv}
\caption{Invariants for state changes}
\label{stateChange}
\end{figure}

\begin{figure}\small
\begin{inv}
\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$
\label{pSendReqPre}
\end{inv}
\begin{inv}
\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$
\label{cSendReqPre}
\end{inv}
\caption{Invariants for sending a request}
\label{sendReq}
\end{figure}

\begin{figure}\small
\begin{inv}
$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{pHandleReq}
\end{inv}
\begin{inv}
$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from $p$, or
\item a thread in $p$ is handling a request for address
$a$ from one of $p$'s children $c'$, and that thread will not send any
(downgrade) request to any child for address $a$ anymore
\end{enumerate}
\label{cHandleReq}
\end{inv}
\caption{Invariants for receiving a request (used by the scheduler)}
\label{recvReq}
\end{figure}

\begin{figure}\small
\begin{inv}
Address $a$ present in node $c$ can not be chosen for eviction by a thread
executing in $c$ if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{evict}
\caption{Invariants for suspending a thread during choosing an address for eviction}
\end{inv}
\end{figure}

\begin{figure}\small
\begin{inv}
Each node has 
\begin{enumerate}
\item at least one dedicated thread resource for handling only messages (requests or
responses) from the parent,
\item at least one dedicated thread resource for handling only requests from any
child, and
\item at least one dedicated thread resource for handling only responses (from children
or from the parent).
\end{enumerate}
\label{dedicate}
\end{inv}
\caption{Dedicated resources in a node}
\end{figure}

\begin{figure}\small
\begin{inv}
Starvation-freedom: A thread that keeps becoming ready to execute will
eventually be scheduled to execute.
\end{inv}
\end{figure}

>>>>

\section{Local Properties}
\label{sec:localProperties}

In this section we list the local properties that are required to ensure the
invariants in Section \ref{sec:properties}. We can easily see by inspection that the distributed MSI protocol in Figure \ref{realistic} does not violate the local properties in this section.

 which in turn makes the 
distributed cache coherence protocol given in Figure \ref{realistic} correct. We
 give the intuition behind the these invariants whenever applicable, though the
formal proof on why these local properties guarantee the invariants of Section
\ref{sec:theorem} is beyond the scope of this paper.

In all the scenarios that we present below, we will use a system containing two
levels of caches, L1 and L2. There are 2 L1 caches $c_1$ and $c_2$, and both are
connected to the shared L2. Both $c_1$ and $c_2$ can handle multiple outstanding
requests from the cores.

@@

Let's illustrate the need for Local-Property \ref{cRespFifo}. Let's say $c_1$ is
in state $M$ and $c_2$ in state $I$ for address $a$, and the corresponding
directory states in L2 match with their actual states. $c_2$ gets a store
request from its processor and sends an upgrade request to L2. L2 sends a
downgrade to $S$ request to $c_1$. $c_1$ receives the downgrade to $S$ request
and downgrades to $S$, sending a downgrade response to L2. Let's say $c_1$
decides to evict address $a$ to make room for another address. $c_1$ sends a
downgrade to $I$ response to L2. Let's say the second response (downgrade to
$I$) reaches L2 first, violating Local-Property \ref{cRespFifo}. L2 sends a
upgrade to $S$ response to $c_2$. Let's say $c_2$ received the upgrade response
and changed state to $S$.  L2 then receives the second downgrade to $S$ response
from $c_1$.  Now L2 will change its directory state to $S$ for $c_1$ because it
received a downgrade to $S$ response last. Now let's say $c_2$ sent an upgrade
to $M$ request to L2 because of a store from its core. L2 will send an downgrade
to $I$ request to $c_1$, but $c_1$ drops the request since it's already in state
$I$.  Hence L2 will never receive a response from $c_1$ leading to a deadlock.
This is illustrated in Figure \ref{complicated}.

Let's illustrate the need for Local-Property \ref{reqNoOvertakeResp}. Let's say
$c_1$ is in state $S$ and $c_2$ in state $I$ for address $a$. Suppose $c_1$
receives a store request and sends an upgrade to $M$ request to L2. L2 sends
back an upgrade to $M$ response to $c_1$. $c_2$ gets a load request for address
$a$ and sends an upgrade to $S$ request to L2. L2 sends a request to $S$ to
$c_1$. Suppose Local-Property \ref{reqNoOvertakeResp} is violated and the
request to $S$ reaches $c_1$ before the response to $M$. $c_1$ will drop the
request since its already in state $S$, and L2 will never get a response leading
to a deadlock. This is illustrated in Figure \ref{lessComplicated}.

\begin{figure}\small
\begin{inv}
A response \Resp{p}{c}{a}{x}, where $p = c.parent$, can be sent from $p$ only
if
\begin{enumerate}
\item a thread that handling a request from $c$ to $p$ for
address $a$ is executing, and
\item $x > y$, where $p.dir[c][a] = y$ just before sending the response
\end{enumerate}
\label{cSendRespPre}
\end{inv}
\begin{inv}
A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
if $x < y$, where $c.state[a] = y$ just before sending the
response\label{pSendRespPre1}
\end{inv}
\begin{inv}
If there is a thread in node $c$ that is waiting to receive a response from its
parent $p$ for an address $a$, then response \Resp{c}{p}{a}{x} can be sent only
on executing a thread that is handling a request from $p$ to $c$ for address
$a$. For instance, $a$ can not be evicted from $c$, if there is a thread
handling a request from $p$ to $c$ for address $a$.
\label{pSendRespPre2}
\end{inv}
\caption{Local-Properties for sending responses}
\label{sendResp}
\end{figure}

We will illustrate why Local-Properties \ref{cSendRespPre} and
\ref{pSendRespPre2} are important. Let's say $c_1$ is in $S$ and $c_2$ is in $I$
for address $a$ the corresponding directory states match with the actual states.

In the first scenario, let $c_1$ receive a store request for address $a$. It
sends an upgrade to $M$ request to L2. Then $c_1$ gets another load request from
the core for a different address $a'$ which is not present in $c_1$. If $c_1$
violates \ref{cSendRespPre} and evicts address $a$ to allocate a line for
address $a$, then it sends a downgrade to $I$ response for address $a$ to L2.
Meanwhile L2 has sent an upgrade to $M$ response to $c_1$. After sending the
response, L2 receives the downgrade to $I$ response form $c_1$. So L2 now
assumes that $c_1$ is in $I$ while in reality, $c_1$ is in $M$. This breaks
Invariant \ref{conservative} which creates the problems described earlier. This
is shown in Figure \ref{checkit}

In the second scenario, let's say L2 violates \ref{pSendRespPre2} and
voluntarily sends an upgrade to $M$ response for address $a$ to $c_1$.
Meanwhile, $c_1$ receives a load request from the core for a different address
$a'$ which is not present in $c_1$, and hence evicts $a$ to allocate a line for
$a$. $c_1$ then gets the upgrade to $M$ response from L2 while L2 gets the
downgrade to $I$ response, breaking Invariant \ref{conservative} again. This is
shown in Figure \ref{checkit2}.

In general, any crossing of response messages between a cache and its parent,
like the ones shown in Figures \ref{checkit} and \ref{checkit2} results in
violation of Invariant \ref{conservative}.

\begin{figure}\small
\begin{inv}
If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$\label{cSendRespPost}
\end{inv}
\begin{inv}
If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$\label{pSendRespPost}
\end{inv}
\begin{inv}
On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
\gets x$\label{pRecvResp}
\end{inv}
\begin{inv}
On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
\gets x$\label{cRecvResp}
\end{inv}
\begin{inv}
$c.state[a]$ can change only on $c$ sending a response to or receiving a
response from $c.parent$\label{cState}
\end{inv}
\begin{inv}
$p.dir[c][a]$ can change only on $p$ sending or receiving a response from
its child $c$\label{pState}
\end{inv}
\caption{Local-Properties for state changes}
\label{stateChange}
\end{figure}

\begin{figure}\small
\begin{inv}
\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$\label{pSendReqPre}
\end{inv}
\begin{inv}
If \Req{p}{c}{a}{x}, where $p = c.parent$, is received, if $x \ge c.state[a]$,
then the request is dropped\label{pSendReqPost}
\end{inv}
\begin{inv}
\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$\label{cSendReqPre}
\end{inv}
\begin{inv}
If \Req{c}{p}{a}{x}, where $p = c.parent$, is received, it $x \le p.state[a]$,
then the request is dropped\label{cSendReqPost}
\end{inv}
\caption{Local-Properties for sending and receiving a request}
\label{sendReq}
\end{figure}

Local-Properties \ref{pSendReqPre} and \ref{pSendReqPost} are needed together
because an incoming request is dropped if the state is already lower than or
equal to the requested state without sending a response. So, the cache waiting
for a response will never receive one, causing a deadlock.  Similarly,
Local-Properties \ref{cSendReqPre} and \ref{cSendReqPost} are also needed
together.

@@

\begin{figure}\small
\begin{inv}
$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{pHandleReq}
\end{inv}
\begin{inv}
$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from $p$, or
\item a thread in $p$ is handling a request for address
$a$ from one of $p$'s children $c'$, and that thread will not send any
(downgrade) request to any child for address $a$ anymore
\end{enumerate}
\label{cHandleReq}
\end{inv}
\caption{Local-Properties for receiving a request (used by the scheduler)}\label{recvReq}
\end{figure}

\begin{figure}\small
\begin{inv}
Address $a$ present in node $c$ can not be chosen for eviction by a thread
executing in $c$ if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from any source
\end{enumerate}\label{evict}
\caption{Local-Properties for suspending a thread during choosing an address for eviction}
\end{inv}
\end{figure}

\begin{figure}\small
\begin{inv}
Each node has at least one dedicated thread resource for handling only messages
(requests or responses) from the parent\label{dedicate1}
\end{inv}
\begin{inv}
Each node has at least one dedicated thread resource for handling only responses
(from children or from the parent)\label{dedicate2}
\end{inv}
\begin{inv}
Each node has at least three thread resources\label{dedicate3}
\end{inv}
\caption{Dedicated resources in a node}
\end{figure}

Local-Properties \ref{evict}, \ref{pHandleReq} and \ref{cHandleReq} are needed
because once a thread has finished up-to a point in its execution, it can not
re-execute that portion. For instance, consider a thread handling an upgrade
request from a child. Let's say thread $t$ has progressed to the point where it
is simply waiting to send an upgrade response to the child. Now, if the line
gets evicted because it was chosen for replacement, or it got downgraded by the
parent, then thread $t$ can not re-execute its procedure again. To prevent this,
only one thread is allowed to operate on a particular address, except in the
case of downgrade requests from the parent. In this case, a thread is created to
handle the downgrade request from the parent even if there is another thread
handling a request for the same address from a child, if the latter thread is
just waiting for a response from the parent. This is done to avoid the deadlock
shown in the following scenario.

Let's illustrate the need for Local-Properties \ref{cReqNoBlockPReq},
\ref{dedicate1}, \ref{cReqNoBlockPReq} and \ref{dedicate2}. Let's say both $c_1$
and $c_2$ are in state $S$ for address $a$, and the corresponding directory
states in L2 match with their actual states. Both $c_1$ and $c_2$ get store
requests from the processor for address $a$, and they send upgrade to $M$
requests to L2. Let's say that L2 received the request from $c_2$ first.  L2
sends a downgrade to $I$ request to $c1$.

Let's first say Local-Property \ref{cReqNoBlockPReq} is not held, and instead some
request from the core is blocking $c_1$ from receiving the request. The $c_1$
will never respond back to L2. But L2 will never start handling the request from
$c_1$ because of Local-Property \ref{cHandleReq}. This will create a deadlock.
The same deadlock is also created when $c_1$ is out of thread resources to
create a thread to handle the downgrade request from $L2$ (violating Local-Property
\ref{dedicate1}).%Figure \ref{deadlockByPBlock} depicts this scenario.

Let's extend this scenario further. Instead of violating Local-Properties
\ref{cReqNoBlockPReq} or \ref{dedicate1} and $c_1$ not responding, let $c_1$
send a downgrade to $I$ response to L2. Let's instead assume that Local-Property
\ref{reqNoBlockResp} is violated in L2. So, the downgrade to $I$ response from
$c_1$ is not received by L2, leading to the same deadlock. In the same
manner, if Local-Property \ref{dedicate2} is violated, the response can not be
handled by L2 creating the same deadlock yet again.

The need for Local-Property \ref{dedicate3} can be shown easily. Let's say that a
cache has exactly 2 thread resources, one for handling messages from parents,
and one for handling responses from either the parent or the children. Now, if
it receives a request from one of its children, it can not use either of the
thread resources. Thus, at least one more thread resource is needed for
handling requests from children. This thread resource can also be used for
handling any type of incoming message.

\begin{figure}\small
\begin{inv}
Starvation-freedom: A thread that keeps becoming ready to execute will
eventually be scheduled to execute.\label{starvation}
\end{inv}
\end{figure}

Local starvation-freedom is required in each cache node. If a node never
executes a particular suspended thread, even if it keeps getting ready to
execute, then the thread will not progress, leading to a deadlock.

### 2012-11-20 paper/isca40format/Theorems.tex
stuff

\section{Theorems}
\label{sec:theorems}

If Invariants \ref{blocking} to \ref{evict} are obeyed, then we have formally
proved that the following theorems hold. The formal proof is beyond the scope of
the paper.

@@

These theorems are needed to guarantee that Figure \ref{realistic} protocol is
indeed correct. If, for instance, a thread wants to send a message in the
output channel of the node and is not able to send one forever, then the thread
can not proceed further. If a thread is waiting for a response (for a request
previously sent), and it never gets the response, the thread can never proceed
further. A thread can also not proceed if it's waiting for being allocated in a
free line but never gets one because all the lines are locked up. These
scenarios create a deadlock.

We will illustrate the need for Theorems \ref{pcSame}, \ref{cpSame} and
\ref{conservative} using a 2-level of caches having two L1 caches $c_1$ and
$c_2$, and one shared L2 cache $p$.

Theorem \ref{pcSame} and \ref{cpSame} are important to make sure that the data
transfers happen correctly. Let $p.dir[c_1][a] = S$, $p.dir[c_2][a] = I$,
$c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
sends \Req{c_1}{p}{a}{M} to $p$. $p$ receives the request, but assumes that
$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Theorem \ref{pcSame}
is violated. So, it does not transfer the data to $c_1$, resulting in $c_1$
waiting for data, and hence deadlock. A dual condition can be constructed
similarly, in which $c_1$ downgrades its state for address $a$ from $S$ to $I$,
and the parent assumes that $c_1$ is in state $M$, and keeps waiting for data.

Theorem \ref{conservative} is important because a parent node consults its
directory to decide the children it needs to send a downgrade request to.  Let
$p.dir[c_1][a] = S$, $c_1.state[a] = S$, $p.dir[c_2][a] = I$ but
$c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
\Req{c_1}{p}{a}{M} to $p$, $p$ will assume that $c_2$'s state for address $a$
is also $I$ and hence sends \Resp{p}{c_1}{a}{M} to $c_1$. But since the real
state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
\ref{singleWriter}.

If Theorems \ref{canSend}, to \ref{cpSame} holds, then the procedures given in
Figure \ref{realistic} can not deadlock because it will always be able to a)
send a request, b) send back a response, c) get back a response for a request it
sent, d) receive data when required. Procedure \uReq{} in Figure
\ref{msi-template} relies on Theorem \ref{conservative} to maintain the
Single-writer Invariant \ref{singleWriter}.

>>>>

\section{Invariants}
\label{sec:properties}

We give a list of invariants or global properties that are needed to ensure that
the distributed, hierarchical, MSI protocol in Figure \ref{realistic} is correct.

@@

The need for these invariants is intuitive. If, for instance, a thread wants to
send a message in the output channel of the node and is not able to send one
forever, then the thread can not proceed further. If a thread is waiting for a
response (for a request previously sent), and it never gets the response, the
thread can never proceed further. A thread can also not proceed if it's waiting
for being allocated in a free line but never gets one because all the lines are
locked up. These scenarios create a deadlock.

We will illustrate the need for Invariants \ref{pcSame}, \ref{cpSame} and
\ref{conservative} using a 2-level of caches having two L1 caches $c_1$ and
$c_2$, and one shared L2 cache $p$.

Invariant \ref{pcSame} and \ref{cpSame} are important to make sure that the data
transfers happen correctly. Let $p.dir[c_1][a] = S$, $p.dir[c_2][a] = I$,
$c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
sends \Req{c_1}{p}{a}{M} to $p$. $p$ receives the request, but assumes that
$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Invariant \ref{pcSame}
is violated. So, it does not transfer the data to $c_1$, resulting in $c_1$
waiting for data, and hence deadlock. A dual condition can be constructed
similarly, in which $c_1$ downgrades its state for address $a$ from $S$ to $I$,
and the parent assumes that $c_1$ is in state $M$, and keeps waiting for data.

Invariant \ref{conservative} is important because a parent node consults its
directory to decide the children it needs to send a downgrade request to.  Let
$p.dir[c_1][a] = S$, $c_1.state[a] = S$, $p.dir[c_2][a] = I$ but
$c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
\Req{c_1}{p}{a}{M} to $p$, $p$ will assume that $c_2$'s state for address $a$
is also $I$ and hence sends \Resp{p}{c_1}{a}{M} to $c_1$. But since the real
state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
\ref{singleWriter}.

If Invariants \ref{canSend}, to \ref{cpSame} holds, then the procedures given in
Figure \ref{realistic} can not deadlock because it will always be able to a)
send a request, b) send back a response, c) get back a response for a request it
sent, d) receive data when required. Procedure \uReq{} in Figure
\ref{msi-template} relies on Invariant \ref{conservative} to maintain the
Single-writer Invariant \ref{singleWriter}.

In order to guarantee these invariants, we need each node to obey certain local
properties that we describe in the next section. Since these properties are
localized to a node, it's easy to see that the distributed, hierarchical, MSI
protocol we derive in Section \ref{sec:DistributedMsi} obeys these properties
just by inspection.
